FraCaS: Temporal Analysis

### 2020-03-11 paper2020/paper.tex
formatting tweaks
tweak
revise

\newlist{subex}{enumerate}{3}
\setlist[subex,1]{parsep=0pt,itemsep=1pt,label=\textbf{P\arabic*}}
\newcommand\hyp{\item[\textbf{H}]}
\newcommand\fracasex[2]{\begin{lingex}\item[(#1)] \begin{subex} #2 \end{subex} \end{lingex} }
\newcommand\jp[1]{\todo[backgroundcolor=blue!10]{JP: #1}}

@@

\begin{enumerate}
\item[exact adverbs] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

  \[⟦\text{at 5 pm, \(s\)}⟧(...) = ⟦s⟧(5pm,5pm)\]

@@

  \[⟦\text{in 1992, \(s\)}⟧(...) = ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)\]

\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include: ``always'', ``never''.

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \[⟦\text{always \(s\)}⟧(t_0,t_1) = ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')\]

  % A particularly interesting case found in FraCas is ``every month'':

  % ⟦``every month s''⟧(t_0,t_1) = ∀t. [t,t+1month] ⊆ [t_0,t_1], ⟦s⟧(t,t+1month)
\end{enumerate}

@@

To be able to conclude as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in P1 and P3) refer to the
same time intervals.

For this purpose we postulate the unicity of action for certain timed
propositions:

\[\constant{unicity}_P : P t_1 t_2 → P t_3 t_4 → (t_1 = t_3) ∧ (t_2 = t_4)\]

P-unicity holds only if P is an \emph{achievement} or an \emph{actitivty}.

@@

A contrario, if P is a \emph{stative}, then we get a time-interval subsumption property:

\[\constant {subsumption}_P : [t_3,t_4] ⊆ [t_1,t_2] → P t_1 t_2 → P t_3 t_4\]

This principle is used for example below:

@@

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:

\paragraph{(un)repeatable achievements}
The principle of using unicity of action interacts well with the usual interpretation of quantifiers (and anaphora).
Indeed, using it, we can refute (279):

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]

Thanks to unicity of action we find \([t_1,t_2] = [t_3,t_4]\), and due
to the years 1991 and 1992 being disjoint we obtain contradiction.

[the scope
for the existential is extended according to our previous analysis of
anaphora]

However, the testsuite instructs that we should \emph{not} be able to refute
(280), with the justification that ``wrote a novel'' is a repeatable
accomplishment:

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, and thus we can't deduce equality of the
interval $t_1,t_2$ and $t_3,t_4$; in turn the hypothesis can't be refuted.

@@

The final class of lexeme carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal in non-trivial ways. Consider for
example ``finish to ...''. The timespan of the argument of finish
should end within the timespan of the finishing action:

\[⟦\text{finish to \(s\)}⟧(t_0,t_1) = ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')\]

@@

The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank three anonymous reviewers for their useful comments on an
earlier draft of the paper.

>>>>

\newlist{subex}{enumerate}{3}
\setlist[subex,1]{parsep=0pt,itemsep=1pt,font=\bfseries,label=P\arabic*}
\newcommand\hyp{\item[H]}
\newcommand\fracasex[2]{\begin{lingex}\item[(#1)] \begin{subex} #2 \end{subex} \end{lingex} }
\newcommand\jp[1]{\todo[backgroundcolor=blue!10]{JP: #1}}

@@

\begin{enumerate}[align=left,font=\itshape]
\item[exact adverbs] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

  \[⟦\text{at 5 pm, \(s\)}⟧(\anyspan) = ⟦s⟧(5pm,5pm)\]

@@

  \begin{multline*}
  ⟦\text{in 1992, \(s\)}⟧(\anyspan) = \\ ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)
\end{multline*}
In the FraCaS test suite, we normally do not find several
time-modifying adverbs modifying a single verb. Indeed, sentences such
as ``in 1992, in 1991 john wrote a novel'' are infelicitous. This
justifies ignoring the input timespan in the above interpretation --
we're in particular not interested in modeling felicity with our
semantics, only giving an accurate semantics when the input is
felicitous.
\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include ``always'' and ``never''.

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \begin{multline*}
  ⟦\text{always \(s\)}⟧(t_0,t_1) = \\ ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')
\end{multline*}
Note that, here, we do use the input interval, allowing to interpret
phrases such as ``In 1994, Itel was always on time.''.

% A specifically interesting case found in FraCas is the adverbial phrase ``every month'', which we modeled
% \begin{equation}
%   ⟦\text{every month s}⟧(t_0,t_1) = ∀t. [t,t+1month] ⊆ [t_0,t_1], ⟦s⟧(t,t+1month)
% \end{equation}
\end{enumerate}

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.

For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:

\[\constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → (t_1 = t_3) ∧ (t_2 = t_4)\]

Unicity of action holds only if the aspectual class of the proposition
$P$ is \emph{achievement} or \emph{actitivty}.\todo{reference in literature}

@@

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid one \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:

\paragraph{Statives}
\textit{A contrario}, if \(P\) is \emph{stative}, then we get a time-interval subsumption property:

\begin{multline*}
\constant {subsumption}_P :\\ [t_3,t_4] ⊆ [t_1,t_2] → P (t_1,t_2) → P (t_3,t_4)
\end{multline*}
This principle is used to reason about problem (314), below:

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because being in Paris is stative, we also get that
Smith was in Paris in any subinterval.

\paragraph{(un)repeatable achievements}
The principle of using unicity of action interacts well with the usual interpretation of quantifiers (and anaphora).
Indeed, using it, we can refute problem (279), as expected in the testsuite:

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Note here that the scope for the existential is extended beyond the
scope of \textbf{P1}. This extension can work according to
\citet{unger:2011}, and our implemented analysis of anaphora\ifanon (citation anonymized) \else\citep{bernardy_computational_2018,bernardy_wide_2019}\fi.

Thanks to the unicity of action of $write(smith,x,...)$ we find \([t_1,t_2] = [t_3,t_4]\), and due
to the years 1991 and 1992 being disjoint we obtain contradiction.

However, the testsuite instructs that we should \emph{not} be able to refute
problem (280), with the justification that ``wrote a novel'' is a repeatable
accomplishment:

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, and thus we cannot deduce equality of the
intervals $t_1,t_2$ and $t_3,t_4$. In turn, the hypothesis cannot be refuted.

@@

The final class of lexemes carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal context in non-trivial
ways. Consider for example ``finish to ...''. The timespan of the
argument of ``finish'' should end within the timespan of the finishing
action:

\begin{multline*}
⟦\text{finish to \(s\)}⟧(t_0,t_1) = \\ ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')
\end{multline*}

@@

The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank anonymous reviewers for their useful comments on an
earlier draft of the paper.
